The Sub-Millimetre User Reduction Facility: an iterative map-maker for SCUBA-2

### 2011-11-25 smurf_paper.tex
Add first part of step correction section

>>>>

\begin{figure*}
\centering
\includegraphics[width=\linewidth]{steps1.pdf}
\caption{Examples of steps in time series data. Steps occur with a wide
range of heights from the large steps shown on the left to the small
steps shown on the right. Large steps are often followed by a brief
``over-shoot'' as shown on the left. In both plots, the black curve is
the uncorrected time series, and the red curve is the corrected time
series. Samples close to a step are omitted in the corrected time series.
In the right hand plot, the blue curve is the uncorrected time series for
a nearby bolometer. The similarity between the red and blue curves shows
that the step correction is performing well.
}
\label{fig:steps1}
\end{figure*}

Sudden steps can occur in the time stream data from each bolometer.
Fig.~\ref{fig:steps1} shows such steps in the time streams for two
bolometers. For each bolometer, correcting these steps involves:

\begin{enumerate}

\item Locating each candidate step.

\item Determining the widths of the candidate steps. Some steps
occur over a single sample such as the step close to sample 5000 in the
right-hand plot of Fig.~\ref{fig:steps1}, but others happen more
gradually such as the step close to sample 5300. In addition, a step can
be preceeded or followed by a short period of instability, as is visible
at the bottom of the step in the right hand plot of
Fig.~\ref{fig:steps1}.

\item Determine the height of the step. Steps can vary enormously in height.
For instance the step in the left plot of Fig.~\ref{fig:steps1} is
roughly 3 pW in height, whereas the step at sample 5100 in the right hand
plot is only 0.01 pW in height. The height of a step is taken as the
difference in the local median value just before and just after the step.

\item Correcting all subsequent data values by the step height.

\item Once all steps have been corrected, a global offset is added to the
corrected bolometer time stream to ensure that the mean value before and
after step correction is unchanged.

\end{enumerate}

\paragraph{Detecting Candidate Steps:}
The first stage in the algorithm is to search the bolometer time stream for
contiguous blocks of samples that have unusually large mean gradients
compared to the neighbouring samples. Each such block of samples is
considered to be a candidate step. The starting and ending time of each
such block is noted. Once all candidate steps have been found for a
bolometer, the algorithm proceeds to the next stage. Candidate steps are
found as follows:

\begin{enumerate}

\item Smooth the bolometer time stream using a median filter of width
50 samples. This reduces the effects of the noise in the data. Using a
median filter rather than a mean filter helps to preserve the sharp edges
at each step, and is robust against spikes, etc.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{steps2.pdf}
\caption{A section of the right hand bolometer from  Fig.~\ref{fig:steps1}
with the median smoothed data overlayed in red.}
\label{fig:steps2}
\end{figure}

The red curve in Fig.~\ref{fig:steps2} shows the median smoothed data for
the right hand bolometer from Fig.~\ref{fig:steps1}.

\item At each time sample, find the difference between the two adjacent
values in the median filtered bolometer data. This is proportional to the
gradient of the median filtered bolometer data. Fig.~\ref{fig:steps3}
shows these differences for the sample bolometer shown in previous
figures. The high gradients at the three step are clearly visible at
indices 5040, 5100 and 5280. In addition, the varying gradient of the
underlying noise-free time series can be seen in the lower level changes
in the difference values.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{steps3.pdf}
\caption{Differences between adjacent values of the median smoothed data
in  Fig.~\ref{fig:steps2}.}
\label{fig:steps3}
\end{figure}

\item Steps are located by looking for difference values that exceed 25
times the RMS of the neighbouring difference value. However, the RMS of
the differences can be heavily affected by the gradient of the underlying
noise-free time series. If there is a strong upward or downward trend in
the time series data, the RMS of the difference values will be higher
potentially resulting in some steps being missed. Therefore to enhance
the reliability of the step detection, an attempt is made to remove these
variations before locating the steps. To do this, the differences are
smoothed using mean filter, and the smoothed differences are removed. The
remaining residual difference values represent the gradient due to steps,
spikes, noise, etc. It is important that the smoothed differences are not
polluted by the large values caused by spikes and steps. So before
smoothing, differences of more than three times the RMS difference (taken
over the whole time stream) are identified and flagged. The RMS is then
re-calculated omitting the flagged values and a further two such
rejection iterations are performed. This flags high gradient values
caused by spikes, jumps, point sources, etc. The difference values close
to a step are often not typical of those in the wider neighbourhood, so
each block of contiguous flagged difference values is doubled in width.
These flagged values are shown as black in Fig.~\ref{fig:steps4}, with
the remaining unflagged values shown in red.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{steps4.pdf}
\caption{The red curve shows the data from Fig.~\ref{fig:steps3} plotted with smaller upper
and lower limits. The blue curve shows the results of smoothing the red
curve. Points omitted from the red curve are shown in black.}
\label{fig:steps4}
\end{figure}

\item The differences for the remaining samples are smoothed
using a mean filter of width 50 samples. The smoothing process fills in small
holes in the array (up to 40 samples), but will not fill in larger holes.
Such larger holes are filled in later by linear interpolation. The
resulting smoothed differences (shown in blue in Fig.~\ref{fig:steps4})
provide an estimate of the local gradient of the underlying noise-free
median-filtered data stream.

\item The smoothed differences are subtracted from the total differences
to get the residual differences caused by noise, spikes, steps, etc.
These are show in black in Fig.~\ref{fig:steps5}.

\item The local RMS of the residual differences at each time sample is
found by smoothing the squared residual differences using a median filter
of width 200 samples, and then taking the square root of the smoothed values.
These are show in green in Fig.~\ref{fig:steps5}.

\begin{figure}
\centering
\includegraphics[width=\linewidth]{steps5.pdf}
\caption{The black curve shows the residuals between the red and blue curves in
Fig.~\ref{fig:steps4}. Missing data is filled in by a combination of
smoothing and linear interpolation. The green curve is the local RMS of the data
over a box width of 200 samples.}
\label{fig:steps5}
\end{figure}

\item Blocks of contiguous samples for which the residual difference exceeds
25 times the local RMS of the residual differences are found. Any two
such blocks that are separated by fewer than 100 samples are merged into
a single block. Blocks are ignored if they fail any of the following
tests:

\begin{enumerate}

\item The width of the block must be less than 200 samples.

\item The total of all the SNR values (the absolute residual difference
divided by the local RMS of the residual differences) in the block must be
at least 25.

\item The total of all SNR values in the block must exceed half the value of
the largest single SNR value in the block.

\end{enumerate}

The second and third tests are intended to prevent spikes being treated as
two very close steps, since such steps will have SNR values of opposite
signs and will thus cancel out.

The data in Fig.~\ref{fig:steps5} results in two blocks being defined -
one between samples 5037 and 5102, and another between samples 5286 and
5296. Each of these blocks define the start and end of a candidate step.

\end{enumerate}
